$Z^\prime$ mass limits and the naturalness of supersymmetry

### 2015-03-06 paper/e6ssm-tuning-paper.tex
More adjustments to phrasings and typos and a few cuts of uneeded or confusing claims.

The structure of this paper is as follows.  In section \ref{sec:model}
we review the models we consider.  In section \ref{sec:ewsb} we
specify the EWSB conditions of the models, with particular focus on
how the $Z^\prime$ mass influences the prediction of $M_Z$.  Then in
section \ref{sec:tuning} we introduce our fine tuning measure and our
approach to evaluating it to obtain the individual sensitivities.  The
numerical procedure we use to obtain our results is given in
\ref{sec:numericalprocedure} and then the results are specified in
section \ref{sec:results}.

@@

$U(1)$ and $E_6$ inspired extensions of the MSSM have been studied
very widely in the literature \cite{Gunion:1989we, Gunion:1992hs,
  Binetruy:1985xm, Ellis:1986yg, Ibanez:1986si, Gunion:1986ky,
  Haber:1986gz, Baer:1987eb, Gunion:1987jd, Grifols:1986vr,
  Ellis:1986ip, Morris:1987fm, Drees:1987tp, Ma:1995xk,
  Suematsu:1997tv, Suematsu:1997qt, Suematsu:1997au, Keith:1996fv,
  Keith:1997zb, Gherghetta:1996yr, Demir:1998dk, Langacker:1998tc,
  Hambye:2000bn, Ma:2000jf} (or for reviews see
\cite{Hewett:1988xc,Langacker:2008yv}).  There has also been a lot of
work recently including investigations of the neutralino sector
\cite{Hesselbach:2001ri, Barger:2005hb, Choi:2006fz,
  Barger:2007nv}; the relic abundance of dark matter
\cite{Kalinowski:2008iq}; GUT scale family symmetries which can
explain the hierarchy of masses in the fermion sector and their
associated mixings \cite{Stech:2008wd}; neutrino physics
\cite{Kang:2004ix}; explanations of the matter-anti-matter asymmetry
of the universe though EW baryogenesis or leptogenesis
\cite{Hambye:2000bn,Ma:2000jf,Kang:2004pp}; dipole moments
\cite{GutierrezRodriguez:2006hb}; anomaly mediated SUSY breaking with
D-term contributions \cite{Asano:2008ju} and the (extended) Higgs
sectors \cite{Daikoku:2000ep,Ham:2008xf}.

@@

The gauge coupling running in the E$_6$SSM at the two-loop level leads
to unification more precisely than in the MSSM \cite{King:2007uj} or,
in slightly modified scenarios, two-step unification can take place
\cite{Howl:2007hq,Howl:2007zi}. If the exotic particles are light in
these models this can open up non-standard decays of the SM--like
Higgs boson \cite{Hall:2010ix,Nevzorov:2013tta,Athron:2014pua}. The
correct relic abundance could be obtained entirely through an almost
decoupled ``inert'' neutralino sector \cite{Hall:2009aj}.  However
this is no longer phenomenologically viable due to limits from direct
detection of dark matter
\cite{2011PhRvL.107m1302A,2012PhRvL.109r1301A, Akerib:2013tjd} and due
to a significant suppression of the decay of the lightest Higgs boson
into SM states, due to a new channel into inert singlinos opening up.

There are still several remaining options. One may specialize to
scenarios known as the EZSSM \cite{Hall:2011zq} where the inert
singlinos that cause these problems are entirely decoupled and the
relic abundance is fitted with a bino-like candidate with a novel
mechanism involving back-scattering into a heavier inert
Higgsino. Another well motivated scenario admits two possible dark
matter candidates \cite{Nevzorov:2012hs}, where one will be an inert
singlino and the other will have a similar composition to MSSM
neutralinos. The simplest phenomenologically viable solution in that
case is to make the singlinos extremely light hot dark matter
candidates and accounts for almost all the relic abundance density with
the lightest ordinary neutralino.

The impact of gauge kinetic mixing in the case where both of the extra
U(1)'s appearing from the breakdown of E$_6$ are present at low energy
was studied in \cite{Rizzo:2012rf}.  The E$_6$SSM was also included in
studies looking at how first or second generation sfermion masses
can be used to constrain the GUT scale parameters \cite{Miller:2012vn}
and the renormalization of VEVs \cite{Sperling:2013eva,
  Sperling:2013xqa}.  The particle spectrum and collider signatures of
the cE$_6$SSM have been studied in a series of
papers,
\cite{Athron:2009ue,Athron:2009bs,Athron:2011wu,Athron:2012sq}. The
threshold corrections to the $\overline{DR}$ gauge and Yukawa
couplings in the E$_6$SSM have also been calculated and their numerical impact in
the constrained version examined \cite{Athron:2012pw}.
%%%%%%%%%%%%%%%%%%---inserted from introduction---%%%%%%%%%%%%%%%%5%%

@@

require that the $Z'$ mass be large, with bounds of $M_{Z'}\gtrsim
2.51$ TeV in $U(1)_\psi$ models and $M_{Z'}\gtrsim 2.62$ TeV in
$U(1)_\chi$ models. Bounds on the $Z^\prime$ mass for other $E_6$ inspired models, such as the $U(1)_N$ of the E$_6$SSM, lie between those on the $U(1)_\psi$ and $U(1)_\chi$. To satisfy these limits typically requires large
values of the singlet VEV $s$. For example $s\gtrsim 6$ TeV is
required in the E$_6$SSM with $U(1)'=U(1)_N$, so that
$|D_{H_1}|,|D_{H_2}|\gg M_Z^2$ for $E_6$ models with $Q_S\neq
0$. As a result the remaining terms on the right-hand side of
Eq. (\ref{eq:E6MZequation}) must be tuned to cancel the resulting very
large contribution to $M_Z$. Moreover, because this is a large tree
level fine tuning, it may negate the increase in naturalness that is
associated with having a reduced need for heavy superpartners. In
$U(1)$ extended models for which $Q_S\neq 0$, the importance
of the $Z^\prime$ mass to the fine tuning in these models can be made
even clearer by writing Eq. (\ref{eq:E6MZequation}) in the form given
in \cite{Athron:2013ipa},

@@

The sensitivity coefficients $\Delta_p$ may be calculated directly

@@

for the $\partial v_i /\partial p$. The sensitivity coefficients

@@

\begin{figure}[h]
\begin{center}
\resizebox{!}{6.1cm}{%
\includegraphics{mssm_stop_tuning.png}}
\resizebox{!}{6.1cm}{%
\includegraphics{mssm_higgs_tuning.png}}
\caption{\textit{Left panel:} Scatter plot of fine tuning in the MSSM as a function of
the lightest stop mass, $m_{\tilde{t}_1}$, for the cutoff scales (from 
bottom to top) $M_X=20$ TeV, $M_X=50$ TeV, $M_X=100$ TeV and $M_X=10^{16}$ GeV. 
\textit{Right panel:} Scatter plot of fine tuning in the MSSM as a function of
the lightest Higgs mass, $m_{h_1}$, for the cutoff scales (from 
bottom to top) $M_X=20$ TeV, $M_X=50$ TeV, $M_X=100$ TeV and $M_X=10^{16}$ GeV.}
\label{Fig:mssmstoptuning}
\end{center}
\end{figure}
To illustrate this, in the left panel of Fig.~\ref{Fig:mssmstoptuning}
we show the variation in fine tuning when we scan over the stop masses
and mixing, with $500 \textrm{ GeV} \leq m_{Q_3},m_{u_3}\leq
10\textrm{ TeV}$ and $-3810\textrm{ GeV}\leq A_t\leq -20\textrm{
  GeV}$, for $M_X=20$ TeV, $50$ TeV and $100$ TeV.  The remaining
parameters we fix such that at $M_{\rm SUSY}$ they have the values
$\mu=-97.5$ GeV, $B=-84.8$ GeV, $M_1=92.1$ GeV, $M_2=95.9$ GeV,
$M_3=352$ GeV, $A_b=-117.9$ GeV, $A_\tau=-7.8$ GeV, $m_{L_i}=400$ GeV,
$m_{e_i}=204$ GeV, $m_{Q_{1,2}}=438$ GeV, $m_{u_{1,2}}436$ GeV and
$m_{d_i}=438$ GeV ($i=1,2,3$). Although we should stress that making
this choice will lead to a spectrum which is in conflict with the LHC
limits, doing so ensures that fine tuning due to the other parameters
is small, so that we avoid washing out the fine tuning impact of the
stops when the tuning is small\footnote{For models in which the
  spectrum is heavier, when the stop masses are small the fine tuning
  reaches a lower bound imposed by other heavier parameters.} as can
be the case when the stop masses are less than 1 TeV.  Note that the
Higgs mass is also allowed to vary in this scan, as shown in the right
panel of Fig.~\ref{Fig:mssmstoptuning}. This illustrates the tuning
problem which people have been worrying about since the discovery of
the 125 GeV Higgs boson as we see that raising the stop masses is also
pushing up the Higgs mass, meaning that heavier Higgs masses require
more fine tuning.  However for a low value of the UV scale this tuning
is not so severe unless the stops are very heavy, and a 125 GeV Higgs
can be obtained without much tuning in this unrealistic case where we
have minimised other sources of tuning.  On the other hand the tuning
becomes more severe as we increase the cutoff such that for $M_X =
10^{16}$ GeV a lightest stop masses of $1-3$ TeV can result in a fine
tuning of $\approx 100 - 1000$ and the minimum tuning we see consistent
with a $125$ GeV Higgs is $\approx 200$, as shown in
Fig.~\ref{Fig:mssmstoptuning}.

@@

Since the stop mass does not have such a large impact on the fine
tuning when the cutoff scale is very low we can use this to see more
clearly the impact of the $Z^\prime$ mass on fine tuning. To do so we
select a fixed low cutoff of $M_X=20$ TeV and compare the fine tuning
for two different values of the $Z^\prime$ mass. We choose to look at
$M_{Z^\prime}=2.5$ TeV, which is just above the current limits, and
$M_{Z^\prime}=4.5$ TeV, which should be in reach in run II at the LHC
\cite{Godfrey:2013eta} and then compare the fine tuning calculated in
each case to the tuning in the MSSM.  For this, we have performed a
large six dimensional parameter space scan in both the MSSM and
E$_6$SSM, varying those parameters most relevant for the fine tuning
and the Higgs mass.  Therefore the set of parameters which we vary
includes $\mu$, $B$ and $\tan \beta$ for the MSSM, and $\lambda$,
$A_\lambda$ and $\tan \beta$, for the E$_6$SSM, which appear at tree
level in the EWSB conditions of the models. While the RGE contribution
from large stop masses to the fine tuning is small for such a low
cutoff scale, the stop contributions to the effective potential can
play a significant role in reducing the fine tuning.  For this reason it is still important to properly treat the
tuning associated with stop contributions to the one-loop effective
potential, and so we also scan over the soft masses $m_{Q_3}^2$,
$m_{u_3}^2$ and the stop mixing $A_t$.  The relevant parameters and
ranges that were scanned over are summarised in Table
\ref{tab:scanranges}. With the exception of the stop mixing $A_t$ and
the soft trilinear $A_\lambda$, which are logarithmically scanned, all
parameters are linearly scanned.  In addition to this we also repeat
each scan for three different values of $M_2$ to allow more variation
in the chargino masses.

@@

MSSM benchmark BM2 (yellow in top panel, individual sensitivities in bottom left panel) was designed to be similar to E$_6$SSM BM2, for a reasonable comparison.  However from the individual sensitivities one can see that the behaviour is quite similar to MSSM BM1, though in this case the $\Delta_{m_{H_u}^2}$ becomes the largest tuning at a higher $M_X$ and does not reach such large values, since more of the tuning is from the mixing in this case.

@@

For a running parameter $q$ appearing in the tree level EWSB conditions, the corresponding contribution to the sensitivity coefficient can then be written 

>>>>

The structure of this paper is as follows.  In section \ref{sec:model}
we review the models we consider.  In section \ref{sec:ewsb} we
specify the EWSB conditions of the models, with particular focus on
how the $Z^\prime$ mass influences the prediction of $M_Z$.  Then in
section \ref{sec:tuningmeasure} we introduce our fine tuning measure
and our approach to evaluating it to obtain the individual
sensitivities.  The results are then given in section
\ref{sec:results}.

@@

$U(1)$ and $E_6$ inspired extensions of the MSSM have been studied
very widely in the literature \cite{Gunion:1989we, Gunion:1992hs,
  Binetruy:1985xm, Ellis:1986yg, Ibanez:1986si, Gunion:1986ky,
  Haber:1986gz, Baer:1987eb, Gunion:1987jd, Grifols:1986vr,
  Ellis:1986ip, Morris:1987fm, Drees:1987tp, Ma:1995xk,
  Suematsu:1997tv, Suematsu:1997qt, Suematsu:1997au, Keith:1996fv,
  Keith:1997zb, Gherghetta:1996yr, Demir:1998dk, Langacker:1998tc,
  Hambye:2000bn, Ma:2000jf} (or for reviews see
\cite{Hewett:1988xc,Langacker:2008yv}).  There has also been a lot of
work recently including investigations of the neutralino sector
\cite{Hesselbach:2001ri, Barger:2005hb, Choi:2006fz,
  Barger:2007nv}; the relic density of dark matter
\cite{Kalinowski:2008iq}; GUT scale family symmetries which can
explain the hierarchy of masses in the fermion sector and their
associated mixings \cite{Stech:2008wd}; neutrino physics
\cite{Kang:2004ix}; explanations of the matter-anti-matter asymmetry
of the universe though EW baryogenesis or leptogenesis
\cite{Hambye:2000bn,Ma:2000jf,Kang:2004pp}; dipole moments
\cite{GutierrezRodriguez:2006hb}; anomaly mediated SUSY breaking with
D-term contributions \cite{Asano:2008ju} and the (extended) Higgs
sectors \cite{Daikoku:2000ep,Ham:2008xf}.

@@

The gauge coupling running in the E$_6$SSM at the two-loop level leads
to unification more precisely than in the MSSM \cite{King:2007uj} or,
in slightly modified scenarios, two-step unification can take place
\cite{Howl:2007hq,Howl:2007zi}. If the exotic particles are light in
these models this can open up non-standard decays of the SM--like
Higgs boson \cite{Hall:2010ix,Nevzorov:2013tta,Athron:2014pua}.

The correct relic density could be obtained entirely through an almost
decoupled ``inert'' neutralino sector \cite{Hall:2009aj}.  However
this is no longer phenomenologically viable due to limits from direct
detection of dark matter
\cite{2011PhRvL.107m1302A,2012PhRvL.109r1301A, Akerib:2013tjd} and due
to a significant suppression of the decay of the lightest Higgs boson
into SM states, due to a new channel into inert singlinos opening up.

There are still several remaining options. One may specialize to
scenarios known as the EZSSM \cite{Hall:2011zq} where the inert
singlinos that cause these problems are entirely decoupled and the
relic abundance is fitted with a bino-like candidate with a novel
mechanism involving back-scattering into a heavier inert
Higgsino. Another well motivated scenario admits two possible dark
matter candidates \cite{Nevzorov:2012hs}, where one will be an inert
singlino and the other will have a similar composition to MSSM
neutralinos. The simplest phenomenologically viable solution in that
case is to make the singlinos extremely light hot dark matter
candidates, in which case the lightest ordinary neutralino accounts
for almost all of the observed relic abundance.

The impact of gauge kinetic mixing in the case where both of the extra
$U(1)$ symmetries appearing from the breakdown of E$_6$ are present at
low energy was studied in \cite{Rizzo:2012rf}.  The E$_6$SSM was also
included in studies looking at how first or second generation sfermion
masses can be used to constrain the GUT scale parameters
\cite{Miller:2012vn} and the renormalization of VEVs
\cite{Sperling:2013eva, Sperling:2013xqa}.  The particle spectrum and
collider signatures of the cE$_6$SSM have been studied in a series of
papers,
\cite{Athron:2009ue,Athron:2009bs,Athron:2011wu,Athron:2012sq}. The
threshold corrections to the $\overline{DR}$ gauge and Yukawa
couplings in the E$_6$SSM have also been calculated and their
numerical impact in the constrained version examined
\cite{Athron:2012pw}.
%%%%%%%%%%%%%%%%%%---inserted from introduction---%%%%%%%%%%%%%%%%5%%

@@

require that the $Z'$ mass be large, with for example bounds of
$M_{Z'}\gtrsim 2.51$ TeV in $U(1)_\psi$ models and $M_{Z'}\gtrsim
2.62$ TeV in $U(1)_\chi$ models. To satisfy these
limits typically requires large values of the singlet VEV $s$. For
example $s\gtrsim 6$ TeV is required in the E$_6$SSM with
$U(1)'=U(1)_N$, so that $|D_{H_1}|,|D_{H_2}|\gg M_Z^2$ for $E_6$
models with $Q_S\neq 0$. As a result the remaining terms on the
right-hand side of Eq. (\ref{eq:E6MZequation}) must be tuned to cancel
the resulting very large contribution to $M_Z$. Moreover, because this
is a large tree level fine tuning, it may negate the increase in
naturalness that is associated with having a reduced need for heavy
superpartners. In $U(1)$ extended models for which $Q_S\neq 0$, the
importance of the $Z^\prime$ mass to the fine tuning in these models
can be made even clearer by writing Eq. (\ref{eq:E6MZequation}) in the
form given in \cite{Athron:2013ipa},

@@

The sensitivity $\Delta_p$ may be calculated directly

@@

for the $\partial v_i /\partial p$. The sensitivities 

@@

\begin{figure}[h]
\begin{center}
\resizebox{!}{6.1cm}{%
\includegraphics{mssm_stop_tuning.png}}
\resizebox{!}{6.1cm}{%
\includegraphics{mssm_higgs_tuning.png}}
\caption{\textit{Left panel:} Scatter plot of fine tuning in the MSSM as a function of
the lightest stop mass, $m_{\tilde{t}_1}$, for the cutoff scales (from 
bottom to top) $M_X=20$ TeV, $M_X=50$ TeV, $M_X=100$ TeV and $M_X=10^{16}$ GeV. 
\textit{Right panel:} Scatter plot of fine tuning in the MSSM as a function of
the lightest Higgs mass, $m_{h_1}$, for the cutoff scales (from 
bottom to top) $M_X=20$ TeV, $M_X=50$ TeV, $M_X=100$ TeV and $M_X=10^{16}$ GeV.}
\label{Fig:mssmstoptuning}
\end{center}
\end{figure}
To illustrate this, in the left panel of Fig.~\ref{Fig:mssmstoptuning}
we show the variation in fine tuning for $M_X=20$ TeV, $50$ TeV, $100$
TeV and $10^{16}$ GeV when we scan over the stop masses and mixing,
with $500 \textrm{ GeV} \leq m_{Q_3},m_{u_3}\leq 10\textrm{ TeV}$ and
$-3810\textrm{ GeV}\leq A_t\leq -20\textrm{ GeV}$.  The remaining
parameters we fix such that at $M_{\rm SUSY}$ they have the values
$\mu=-97.5$ GeV, $B=-84.8$ GeV, $M_1=92.1$ GeV, $M_2=95.9$ GeV,
$M_3=352$ GeV, $A_b=-117.9$ GeV, $A_\tau=-7.8$ GeV, $m_{L_i}=400$ GeV,
$m_{e_i}=204$ GeV, $m_{Q_{1,2}}=438$ GeV, $m_{u_{1,2}}436$ GeV and
$m_{d_i}=438$ GeV ($i=1,2,3$). Although we should stress that making
this choice will lead to a spectrum which is in conflict with the LHC
limits, doing so ensures that fine tuning due to the other parameters
is small, so that we avoid washing out the fine tuning impact of the
stops when the tuning is small\footnote{For models in which the
  spectrum is heavier, when the stop masses are small the fine tuning
  reaches a lower bound imposed by other heavier parameters.} as can
be the case when the stop masses are less than 1 TeV.  Note that the
Higgs mass is also allowed to vary in this scan, as shown in the right
panel of Fig.~\ref{Fig:mssmstoptuning}. This illustrates the tuning
problem which people have been worrying about since the discovery of
the 125 GeV Higgs boson as we see that raising the stop masses is also
pushing up the Higgs mass, meaning that heavier Higgs masses require
more fine tuning.  However for a low value of the UV scale this tuning
is not so severe unless the stops are very heavy, and a 125 GeV Higgs
can be obtained without much tuning in this unrealistic case where we
have minimised other sources of tuning.  On the other hand the tuning
becomes more severe as we increase the cutoff such that for $M_X =
10^{16}$ GeV a lightest stop masses of $1-3$ TeV can result in a fine
tuning of $\approx 100 - 1000$ and the minimum tuning we see
consistent with a $125$ GeV Higgs is $\approx 200$, as shown in
Fig.~\ref{Fig:mssmstoptuning}.

@@

Since the stop mass does not have such a large impact on the fine
tuning when the cutoff scale is very low we can use this to see more
clearly the impact of the $Z^\prime$ mass on fine tuning. To do so we
select a fixed low cutoff of $M_X=20$ TeV and compare the fine tuning
between the MSSM and E$_6$SSM for two different values of the
$Z^\prime$ mass. We choose to look at $M_{Z^\prime}=2.5$ TeV, which is
just above the current limits, and $M_{Z^\prime}=4.5$ TeV, which
should be in reach in run II at the LHC \cite{Godfrey:2013eta} and
then compare the fine tuning calculated in each case to the tuning in
the MSSM.  For this, we have performed a six dimensional parameter
space scan in both the MSSM and E$_6$SSM, varying those parameters
most relevant for the fine tuning and the Higgs mass.  Therefore the
set of parameters which we vary includes $\mu$, $B$ and $\tan \beta$
for the MSSM, and $\lambda$, $A_\lambda$ and $\tan \beta$, for the
E$_6$SSM, which appear at tree level in the EWSB conditions of the
models. While the RGE contribution from large stop masses to the fine
tuning is small for such a low cutoff scale, the stop contributions to
the effective potential can play a significant role in reducing the
fine tuning.  For this reason it is still important to properly treat
the tuning associated with stop contributions to the one-loop
effective potential, and so we also scan over the soft masses
$m_{Q_3}^2$, $m_{u_3}^2$ and the stop mixing $A_t$.  The relevant
parameters and ranges that were scanned over are summarised in Table
\ref{tab:scanranges}.  In addition to this we also repeat each scan
for three different values of $M_2$ to allow more variation in the
chargino masses.

@@

MSSM benchmark BM2 (yellow in top panel, individual sensitivities in bottom left panel) is designed to be similar to E$_6$SSM BM2, for a reasonable comparison.  However from the individual sensitivities one can see that the behaviour is quite similar to MSSM BM1, though in this case the $\Delta_{m_{H_u}^2}$ becomes the largest tuning at a higher $M_X$ and does not reach such large values, since more of the tuning is from the mixing in this case.

@@

For a running parameter $q$ appearing in the tree level EWSB conditions, the corresponding contribution to the individual sensitivity can then be written 
